Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2021-10-15 doc/JFLA2022/paper.tex
small intro pass

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater than 1 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend on the considered variant of the game
  and the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

Kirby and Paris prove the following theorems, applying
combinatorial results about ordinal numbers by Jussi Ketonen and Robert Solovay~\cite{KS81}.

@@

Complex formalisations and proofs are explained in the book.
% an  electronic book~\cite{HydraBook} (PDF document of over 280 pages).
Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$. The first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}.

This work is also an opportunity to 
 provide concrete examples of formalization and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc. It may be also used as a library on ordinal numbers, for instance for proving termination properties.

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the \community project~\cite{CoqCommunity}, and documentation generated with \alectr~\cite{alectryonpaper, alectryongithub}.

>>>>

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Figure~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater than 1 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend on the considered variant of the game
  and the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

Kirby and Paris proved the following theorems, applying
combinatorial results about ordinal numbers by Jussi Ketonen and Robert Solovay~\cite{KS81}.

@@

Complex formalizations and proofs are explained in the book.
% an  electronic book~\cite{HydraBook} (PDF document of over 280 pages).
Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$. The first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}.

This work is also an opportunity to 
 provide concrete examples of formalization and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc. It may be also used as a library on ordinal numbers, for instance for proving termination properties.
%
 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
In this paper, we present recent evolutions of the library: new results, interaction with the \community project~\cite{CoqCommunity}, and documentation generated with \alectr~\cite{alectryonpaper, alectryongithub}.
